A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-03-31 response/response.tex

    \item
        Pg. 19: Are the units of the vertical axis in Fig. 5 “seconds”, as in Figs. 1 and 2?
        Please indicate that. Also, is the label of the vertical axis in Fig. 6 correct?

{\color{red}  
\begin{itemize}
     \item
        The manuscript has been corrected. The labels on the vertical axis in Fig. 5 were incorrect. 
  \end{itemize}}

\item
Pg. 20: In lines 32-35, clarify what “byte-copyable” mean and why is that important.

{\color{red}  
\begin{itemize}
     \item
     Clarification made in manuscript, new reference added. 
     %
     'Byte-copiable' means that members of an array of data elements are all of
     the same size in memory so that if one knew the position of the first
     element in the array, the position of all other elements in the array
     could be inferred with the additional information of the length of the
     array. This is important, in particular, to the designers and users of the
     thousands of MIC accelerator cards installed in the \emph{Stampede}
     computer system. MIC accelerator cards permit only the transfer of
     byte-copiable data from the host computer system for use with the 'pragma
     offload' mode, which is to say most code written for the MIC. This is even
     more important when it is realized that the great majority of
     \emph{Stampede}'s theoretical processing power, the figure used to advertise to potential clients, is counted from these MIC
     accelerator cards. Meaning that if an AD user wanted to use any of that
     power, they would be out of luck since not even the full time personel at
     TACC could manage to compile the AD library Trilinos::Sacado for the MIC.
     A more 'casual' computational scientist would find the task impossible.
  \end{itemize}}

### 2014-04-08 jacobian_calc_CMAME.tex
fixed some errors in a table
Update Lit. Rev.

%Package for adding notes/comments/and tracking changes
\usepackage[inline]{trackchanges}
\addeditor{JTF}
\addeditor{Michael}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

In order to maintain the quadratic convergence properties of Newton's method in
quasi-static nonlinear analysis of solid structures it is crucial to obtain
accurate, algorithmically consistent tangent-stiffness matrices. A goal of the
study described in this paper was to establish the suitability of an
underexplored method for numerical computation of tangent-stiffness operators,
referred to as ``complex-step'', and compare the \note[Michael]{re: I.1}
\remove[Michael]{new} method with other techniques for numerical derivative
calculation: automatic differentiation, forward finite difference, and central
finite difference. The complex-step method was \note[Michael]{re: I.1, could
suggest claim of novelty} \remove[Michael]{newly} implemented in a massively
parallel computational peridynamics code for the purpose of this comparison.
The methods were compared through profiling of the code for accuracy,
speed, efficiency, and parallel scalability. The research provides data that
can serve as practical guide for code developers and analysts faced with
choosing which method best suits the needs of their application code.
Additionally, motivated by the reproducible research movement, all of the
code, examples, and workflow to regenerate the data and figures in this paper
are provided as open source.

@@

In order to maintain the quadratic convergence properties of the first-order
Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in
quasi-static nonlinear analysis of solid structures it is crucial to obtain
accurate, algorithmically consistent tangent-stiffness matrices. For an
extremely small class of nonlinear material models, these consistent
tangent-stiffness operators can be derived analytically; however, most often in
practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \note[Michael]{re: I.1}
\change[Michael]{new}{relatively underexplored}, accurate, and practical method
for calculating tangent-stiffness matrices against established methods.  This
new method is based on a complex number Taylor series expansion and referred to
as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is
determined by comparing its measured accuracy\note[Michael]{re:
I.1}\remove[Michael]{new} relative to the measured accuracy of popular finite
differencing techniques used for computing tangent-stiffness matrices. For this
comparison the exact to machine precision derivatives computed with
\emph{automatic differentiation} serve as the standard. \note[Michael]{re:
II.1}For the reader's convenience, from plasticity theory, the algorithmic
consistency of the tangent-stiffness matrix refers to consistency with the
integration of a plasticity model under the Khun-Tucker constraints or similar
where the tangent-stiffness matrix is computed using the elastoplastic modulii
that are determined from the integration \cite{simo1998}.

Comparative data regarding the accuracy and compute cycle timing for the
complex step, forward difference, central difference, and automatic
differentiation methods are included to aid developers and analysts in
computational mechanics code design who are faced with implementing a general
method for tangent-stiffness matrix calculation.  A comparison of the methods was achieved through instrumentation of a
massively parallel computational mechanics code.

The scope of the application component of this study was limited to a single
material model, an elastic peridynamic solid, implemented in the computational
peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific
application provided a practical framework for implementing the
\note[Michael]{re: I.1}\remove[Michael]{new} method and solving engineering
problems to generate the data needed to compare the methods. In particular,
\emph{Peridigm} was chosen because it combined several helpful
characteristics: the utilization of Newton's method and tangent-stiffness
matrices for solving nonlinear quasi-static problems, prior inclusion of
forward difference, central difference, and automatic differentiation
methods needed for comparison to the new complex-step method
implementation, and being an agile-components code that makes use of
distributed computing data structures via Trilinos \cite{trilinos} for
efficient parallelization on large clusters. Agile-components is a software
development term meaning to structure a software project for continuous
development to respond to evolving specifications by making much use of
techniques from object-oriented programming and reusing proven pre-existing
code to reduce development time and increase reliability. The modularity of
\emph{Peridigm} allowed the adding of new features, such as those required by the
study.

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing the complex-step method in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic differentiation or complex-step to better encourage their use and
discussion. \note[Michael]{re: II.5}This open source resource is available at the corresponding author's website or at \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

This section contains background information on the differentiation techniques
underlying the tangent-stiffness matrix calculation methods compared in the
study. Since the derivation of first-order finite difference techniques from
Taylor series expansion is considered well known, it is not described here.
Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

@@

It is
possible to approximate derivatives quite accurately with a technique based on
a complex variable Taylor series expansion of a function.  This method was
first described by Lyness and Moler
\cite{lyness1967numerical,lyness1968differentiation} and has more recently been
rediscovered for use in engineering analysis.  The basic idea is that a model parameter can
be made complex and expanded in a Taylor series about a small perturbation
along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i
h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} +
\frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} \end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor})
and solving for the first derivative and ignoring terms of $\mathcal{O}\left(
h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im}
\left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  \label{eqn:complexFirstDeriv} \end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only
utilizing one functional evaluation of a perturbed model parameter along the
imaginary axis.  The step-size $h$ is arbitrary and can be made as small as
practical (even to machine precision without the dangers of roundoff error as
in finite differences) to yield an accurate estimate of this derivative.  The
only disadvantage of this technique is the necessity of requiring the functions
to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a
simple example that illustrates the use of the complex-step method for the
derivative calculation of a function. 
%
\note[Michael]{re: II.6}
For the reader's convenience the forward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

\note[Michael]{re: I.4, John all the rest of this section is new}
It is appropriate to detail the prior work discussing aspects of the
complex-step method in order to indicate what is new in the instant work.  The
CS method was introduced in a work by J. N. Lyness and C. B. Moler
\cite{lyness1967numerical}.  In this prior work, the CS method is identified in
an infinite series form and the computing of analytical derivatives is shown. This prior
work develops the CS method, while the instant work is about an application of
the CS method.

A follow-on work by J. N. Lyness demonstrates a 'truncated' approximate version
of the original CS method to make it suitable for use on digital computers \cite{lyness1968differentiation}.
This form of the CS method is used in the instant work.
The prior work develops the CS method, while the instant work is about an application of the CS method.

A work by W. Squire and G. Trapp shows through empirical data that CS offers superior accuracy
to FD and CD \cite{squire1998using}. The instant work uses the CS method as shown in the
prior work and was motivated by the prior work's suggestion that CS is highly accurate. 
The instant work examines applications of the CS method beyond taking derivatives alone,
whereas the prior work does not.

A work by J. C. Newman, W. K. Anderson and L. D. Whitfield discusses computing
sensitivities in a CFD analysis using complex step and compares the results to
central difference produced results \cite{newman1998}. The work is also notable
for identifying automatic differentiation as a competing method to CS and
suggesting that complex step could be used to compute Jacobians of residuals
for use in Newton-Krylov schemes for solving non-linear systems. The instant
work is novel with respect to the prior work because the instant work
investigates CS as applied to a numerical method for solving non-linear
systems, rather than as in the prior work investigating the derivatives
themselves produced with the CS method.

In a work by A. Perez-Foguet, A. Rodriguez-Ferran, A. Huerta, CS is used to
compute Jacobians for integrating plastic stress to determine
elastoplastic modulii and for computing the Jacobian of the residual for the global
problem, both of which are an application of the CS method to the solution of
non-linear systems \cite{perez2000numerical}. These elastoplastic modulii are
used in computing the global Jacobian  and are said to be 'algorithmically
consistent' in that they are consistent with the algorithm used to return the
predicted stress to the yield surface. Examples of interest include FEM
simulations on two plastic material models with complicated hardening rules.
Comparing the prior work with the instant work, the prior work demonstrates CS
on a serial computer and discusses a classical solid mechanics formulation.
Automatic differentiation is not investigated and CPU time measurements are not
taken. Scalability studies with a variety of equivalent, progressively more
refined meshes are not done. Scalability as a function of parallel computer
size is not investigated. The instant work is novel with respect to the prior work in
offering a discussion of CS as applied to peridynamics simulation on a parallel
computer, CPU time data, and comparison of CS to automatic differentiation as a competing method.

Another work by A. Perez-Foguet, A. Rodriguez-Ferran, A. Huerta applies CS in
computing derivatives needed for integrating stress, obtaining algorithmically
consistent elastoplastic modulii, and obtaining the Jacobian of the global
residual, for FEM simulations using the MRS-Lade material model
\cite{perez2012numerical}. In this prior work, CPU runtime is measured.
Comparing the prior work to the instant work, the prior work demonstrates CS on
a serial computer and discusses a classical solid mechanics formulation.
Automatic differentiation is not investigated.  Scalability studies with a
variety of equivalent, progressively more refined meshes are not done.
Scalability as a function of distributed system size is not investigated.  The
instant work is novel with respect to the prior work in offering a discussion
of CS as applied to peridynamics simulation on a parallel computer, and
comparison of CS to automatic differentiation as a competing method.

A work by Joaquim R. R. A. Martins, Ilan M. Kroo and Juan J. Alonso
investigates using CS for computing aerodynamic sensitivities
\cite{martins2000automated}. Additionally, higher order accuracy versions of CS
are discussed, as is the task of implementing complex-step in most widely used
general purpose programming languages. Also discussed is the relative ease of
implementing a CS scheme versus an AD based scheme for computing sensitivities
in an established code. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the use of the CS method to obtain sensitivities.

A work by W. K. Anderson, J. C. Newman, D. l. Whitfield and E. J. Nielsen
discusses the CS method as applied to computing sensitivities in CFD simulations
\cite{anderson2001sensitivity}. CS is compared to AD both in Fortran and C++
applications. In this prior work it is suggested that forward mode AD is
mathematically equivalent to first order CS. The instant work is novel with
respect to the prior work because the instant work investigates CS as applied
to a numerical method for solving non-linear systems, rather than as in the
prior work investigating the derivatives themselves produced with the CS
method.

A work by J. Martins, P. Sturdza, J. Alonso demonstrates that the CS method can
be performed by using complex datatypes for the variables of functions
comprising an analysis code \cite{martins2003complex}. The analysis code can
comprise a simulation and derivatives of response variables called
sensitivities can be computed. The accuracy of this approach is compared to
automatic differentiation. The instant work is novel with respect to the prior
work because the instant work investigates CS as applied to a numerical method
for solving non-linear systems, rather than as in the prior work investigating
the derivatives themselves produced with the CS method.

A work by  K. L. Lai and J. L. Crassidis discusses first and second order CS
methods for computing derivatives and choosing optimal step size
\cite{lai2008extensions}. The instant work discusses using the derivatives
computed with the CS method in a numerical method for solving a non-linear
system, while in the prior work the derivatives obtained from the CS method 
are the object of interest.

A work by A. H. Al-Mohy and N. J. Higham investigates the use of complex step
in computing Fr\'{e}chet derivatives of matrix to matrix functions
\cite{al2010complex}. To compute the Jacobian of the residual in the
peridynamics simulation as discussed in the instant work is analogous is
computing a Fr\'{e}chet derivative. Where the application of the CS derivatives in
the prior work is limited to a particular class of abstract functions to
determine the CS method's suitability for use in condition number estimation, the
instant work applies CS to solving a solid mechanics problem. The instant work is not
about computing Fr\'{e}chet derivatives with CS but moreover using them in a
numerical method to solve a set of equations describing a physical system.

A work by W. Jin, B. H. Dennis, and B. P. Wang discusses using CS to compute
sensitivities in a solid mechanics FEM analysis \cite{jin2010improved}. In the
prior work the derivatives produced by CS are the object of
interest, whereas the instant work uses CS in a numerical method for solving a
non-linear system.

A work by A. Voorhees, H. Millwater, and R. Bagley discusses the background of
the CS method and frames it as a subset of Fourier Differentiation. CS and
Fourier Differentiation are then applied to computing sensitivities within a
solid mechanics FEA \cite{voorhees2011complex}. A notable technique
demonstrated by this prior work is the representation of complex valued nodal
coordinates with additional degrees of freedom in the FE model, to be
interpreted by a matrix representation of complex numbers during computations.
In the prior work, CS is discussed and related to a broader class of
differentiation methods and used to compute sensitivities. In the instant work,
CS is used in a numerical method for solving non-linear systems.

In another work by H. Millwater, C. Quintana, J. Garza and David Wagner the
concept of incorporating complex values into the degrees of freedom of finite
elements is called ZFEM \cite{millwater2013application}. Notable in this work
is the discussion of creating user defined element types for use with the
popular \emph{ABAQUS} software package which enables the use of ZFEM in that
software. \emph{ABAQUS} is a trademark of Dassault Systemes
\cite{systemes2012abaqus}. In this prior work the goal of using CS through the
ZFEM technique is to compute shape sensitivities, while in the instant work CS
is used in a numerical method for solving non-linear systems.

A work by R. Abreu, D. Stich, and J. Morales discusses computing derivatives
with a version of the CS method which features the combined use
of real and imaginary valued step components \cite{abreu2013generalization}.
This prior work suggests that the modified CS method can achieve extended
approximation accuracy up to fourth order.  The prior work discusses using the
CS method to compute derivatives and attempts to extend the method for that
application, where the derivatives themselves are the object of interest. In
contrast, the instant work discusses using the derivatives computed with CS in
a numerical method for solving a non-linear system.

The body of prior literature has introduced the application of CS to
single dimension and multidimensional derivatives, sensitivity analysis,
non-linear solution schemes, the extension of CS for higher accuracy, 
and implementation strategies for CS in new and existing codes.
With respect to the prior work discussing the CS method, the novel
aspects of the instant work are then its discussion of the application of CS to
peridynamics simulations run in \emph{Peridigm}, a parallel implementation of
CS and a compilation of CPU time data, where each of CS, AD, FD and CD are
compared.

@@

An AD system evaluates composition functions in the expected manner: it works
by first evaluating the innermost function of the composition, then presenting
that function's output as input to the next level function until all levels are
complete, observing commonly expected order of operations. However, an AD
system does additional work during a function evaluation in that as each nested
function is evaluated, the function's partial derivative with respect to the
designated variables of the given input is also calculated.  This is possible
because the elementary math functions are hard-coded into the AD source-code
along with their analytical derivatives, and linked by special instructions, so
that when the elementary math functions are called upon for computation, their
partial derivatives may be computed and stored in a sequence. The AD system
then multiplies the final sequence of partial derivatives together to produce
the exact equivalent to taking a partial derivative of the corresponding
composition function with respect to a designated variable at a particular
value. 
%
It is obvious, but bears mentioning, that the AD system could simply
store one value for partial derivatives, modifying it as appropriate for every
function evaluation rather than keeping a sequence. This is significant because
modifying a single value rather than keeping a list of values corresponding to
every level of a composition function, which itself may not be the sole one
needing to be evaluated, represents a savings in memory usage, which at large
scales is an active concern. In the literature, the AD scheme described here is
called \emph{forward automatic differentiation}. Only forward AD will be
covered here since it is the implementation used in \emph{Sacado} and therefore
in this study; however, the reader is referred to the introduction section of
\cite{ref-AD-methods} and its citations for further information on AD, and
particularly \cite{ref-on-AD} which is foundational.
Appendix~\ref{sec:appendixB} includes an example that walks the reader through
the process a computer uses to compute derivatives via AD.  Some things to note
about AD are that no approximation of derivatives is being made because the
analytical forms of the partial derivatives of the elementary math functions are defined
alongside them. The accuracy of AD is then limited by the precision of the AD
system's definition of the elementary math functions and their partial
derivatives.

@@

\note[Michael]{re: I.5, next section is taken from the thesis.}
\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}

@@

\note[Michael]{re: I.2, John I think you would better handle expanding this section,
I was going to essentially copy silling:psa.}\change[Michael]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom}.

>>>>

%Package for adding notes/comments/and tracking changes
\usepackage[inline]{trackchanges}
\addeditor{JTF}
\addeditor{MDB}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

In order to maintain the quadratic convergence properties of Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. A goal of the study described in this paper was to establish the suitability of an underexplored method for numerical computation of tangent-stiffness operators, referred to as ``complex-step'', and compare the \note{Re: I.1} \remove{new} method with other techniques for numerical derivative calculation: automatic differentiation, forward finite difference, and central finite difference. The complex-step method was \note{Re: I.1} \remove{newly} implemented in a massively parallel computational peridynamics code for the purpose of this comparison.  The methods were compared through profiling of the code for accuracy, speed, efficiency, and parallel scalability. The research provides data that can serve as practical guide for code developers and analysts faced with choosing which method best suits the needs of their application code.  Additionally, motivated by the reproducible research movement, all of the code, examples, and workflow to regenerate the data and figures in this paper are provided as open source.

@@

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent\footnote{\note{Re: II.1} In the context of plasticity thoery, the algorithmic consistency of the tangent-stiffness matrix refers to consistency with the integration of a plasticity model under the Khun-Tucker constraints or similar where the tangent-stiffness matrix is computed using the elastoplastic modulii that are determined from the integration \cite{simo1998}.} tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

A goal of this study was to develop and evaluate a \note{Re: I.1} \change{new }{relatively unexplored}, accurate, and practical method for calculating tangent-stiffness matrices against established methods.  This \note{Re: I.1} \remove{new} method is based on a complex number Taylor series expansion and referred to as CTSE or the ``complex-step'' method. The practical accuracy of CTSE is determined by comparing its measured accuracy relative to the measured accuracy of popular finite differencing techniques used for computing tangent-stiffness matrices. For this comparison the,  exact to machine precision, derivatives computed with \emph{automatic differentiation} serve as the standard.

Comparative data regarding the accuracy and compute cycle timing for the
complex step, forward difference, central difference, and automatic differentiation methods are included to aid developers and analysts in computational mechanics code design who are faced with implementing a general
method for tangent-stiffness matrix calculation.  A comparison of the methods was achieved through instrumentation of a massively-parallel computational mechanics code.

The scope of the application component of this study was limited to a single material model, an elastic peridynamic solid, implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the \note{Re: I.1} \change{new }{complex-step} method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of forward-difference, central-difference, and automatic-differentiation methods needed for comparison to the complex-step method implementation, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters. Agile-components is a software development term meaning to structure a software project for continuous development to respond to evolving specifications by making much use of techniques from object-oriented programming and reusing proven pre-existing code to reduce development time and increase reliability. The modularity of \emph{Peridigm} allowed the adding of new features, such as those required by the study.

The aim of this paper is not only to \note{Re: I.1} \change{introduce the new}{study the} complex-step method in the context of evaluating tangent-stiffness matrices, but to serve as a review for other differentiation techniques useful for solving non-linear systems with Newton's method. After presenting background information on the underlying methods to be discussed, presented in this paper are: In Section~\ref{subsec:TS} a description of tangent-stiffness matrices, detailed directions for producing tangent-stiffness matrices with each of the methods identified above, and a description and justification of the \note{Re: I.1} \remove{new} complex-step method for calculating tangent-stiffness matrices, in Section~\ref{subsubsec:Impl} a description of implementing the complex-step method in the \emph{Peridigm} software, in Section~\ref{JGAM} a description of the quantities-of-interest used to rank the methods, in Section~\ref{sec:Res} a presentation and analysis of the results of the comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.  Finally, in the interest of replicable research, the reader is referred to the corresponding author's website for \emph{Peridigm} C++ source-code and data necessary to reproduce the results presented here. Additionally, included on the website is an example serial C++ library with classes for solving non-linear systems using the techniques discussed here, as well as two validation and verification example problems which show how to use the library.  The purpose of the serial C++ library is to demonstrate in a less complex manner than in the \emph{Peridigm} code how one would use automatic-differentiation or complex-step to better encourage their use and discussion. \note{Re: II.5} \add{This open source resource is available at the corresponding author's website or at} \href{https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive}{https://github.com/utsa-idl/Complex\_Step\_Jacobian\_Project\_Archive}.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in the study. Since the derivation of first-order finite difference techniques from Taylor series expansion is considered well-known, it is not described here.  Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

@@

It is possible to approximate derivatives quite accurately with a technique based on a complex-variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical,lyness1968differentiation} and has more recently been rediscovered for use in engineering analysis.  The basic idea is that a model parameter can be made complex and expanded in a Taylor series about a small perturbation along the imaginary axis by some arbitrary value $h$ as follows
%
\begin{align} 
    f ( x + i h ) =& f(x) + \frac{\partial f(x)}{\partial x} \frac{i h}{1!}  \notag \\ &+ \frac{\partial^2 f(x)}{\partial x^2} \frac{(i h)^2}{2!} + \frac{\partial^3 f(x)}{\partial x^3} \frac{(i h)^3}{3!} + \cdots.
\label{eqn:complexTaylor} 
\end{align}
%
Taking the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solving for the first derivative and ignoring terms of $\mathcal{O}\left(h^2 \right)$ yields an estimate
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{\mbox{Im} \left( f (x + i h) \right)}{h} +\mathcal{O}\left( h^2 \right).  
\label{eqn:complexFirstDeriv} 
\end{equation}
%
Thus, an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter along the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for the derivative calculation of a function. 
%
\note{Re: II.6} \add{For completeness, the forward- and central-difference formulas corresponding to equation}~(\ref{eqn:complexFirstDeriv}) \add{are} \cite[Chap. 4.1.3]{chapra2010}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left( h \right), 
\label{eqn:forwardFirstDeriv} 
\end{equation}
%
\begin{equation} 
\frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left( h^2 \right). 
\label{eqn:centeredFirstDeriv} 
\end{equation} 
%
\add{The order of the truncation error in the forward-difference formula (equation}~(\ref{eqn:forwardFirstDeriv})\add{) indicates that for a given $h$ this method is less accurate an approximation than either central-difference or complex-step which have errors of $\mathcal{O}(h^2)$.}

\note{Re: I.4} \add{As mentioned earlier, the CS method was originally introduced in a paper by Lyness and Moler} \cite{lyness1967numerical}\add{.  A follow-on contribution by Lyness demonstrated a ``truncated'' approximate version of the original CS method to make it suitable for use on digital computers} \cite{lyness1968differentiation}\add{.  After these original papers, little can be found in the literature regarding the method until Squire and Trapp showed through empirical data that CS offers superior accuracy to FD and CD} \cite{squire1998using} \add{in computing numerical derivatives. Papers by Newman et al.\ } \cite{newman1998} \add{and Anderson et al.\ } \cite{anderson2001sensitivity} \add{discuss computing sensitivities in a CFD analysis using CS and compares the results to sensitivities produced with central-difference derivatives. The work is also notable for identifying automatic-differentiation as a competing method to CS and suggesting that complex-step could be used to compute Jacobians of residuals for use in Newton-Krylov schemes for solving non-linear systems. Perez-Foguet et al.\ } \cite{perez2000numerical, perez2012numerical} \add{used the CS method to compute algorithmically consistent elastoplastic moduli in non-linear solid mechanics problems.  A paper by Joaquim et al.\ } \cite{martins2000automated} \add{investigates using CS for computing aerodynamic sensitivities. Additionally, higher-order accuracy versions of CS are discussed, as is the task of implementation in most widely used general purpose programming languages. Discussed is the relative ease of implementing a CS scheme versus an automatic-differentiation based scheme for computing sensitivities in an established code.  
Martins et al.\ } \cite{martins2003complex}  \add{demonstrates that the CS method can be performed by using complex datatypes for the variables of functions comprising an analysis code. The accuracy of this approach, with respect to the derivatives themselves, is compared to automatic-differentiation in a serial analysis code. Lai and Crassidis} \cite{lai2008extensions} \add{discuss first- and second-order CS methods for computing derivatives and choosing optimal step sizes.  Al-Mohy and Higham} \cite{al2010complex} \add{investigate the use of CS in computing Fr\'{e}chet derivatives of matrix functions. Jin et al.\ } \cite{jin2010improved} \add{discuss using CS to compute sensitivities in a solid mechanics FEM analysis.  Voorhees et al.\ } \cite{voorhees2011complex} \add{discusses the background of the CS method and frames it as a subset of Fourier Differentiation. CS and Fourier Differentiation are then applied to computing sensitivities within a solid mechanics finite element analysis. They implemented the complex-values as additional degrees of freedom in the finite element stiffness matrices using a real-valued matrix representation of the complex numbers where required. In a follow-on work,  Millwater et al.\ } \cite{millwater2013application} \add{extended this concept, calling it ZFEM, and implemented the method into user defined element types for use with the popular ABAQUS} \cite{systemes2012abaqus} \add{FEA software package.  Finally, Abreu et al.\ } \cite{abreu2013generalization} \add{discusses computing derivatives with a version of the CS method which features the combined use of real and imaginary valued step components, as in Fourier differentiation.  They suggests that the modified CS method can achieve extended approximation accuracy up to fourth order.  The body of prior literature has introduced the application of CS to single dimension and multidimensional derivatives, sensitivity analysis, non-linear solution schemes, the extension of CS for higher accuracy, and implementation strategies for CS in new and existing codes.  With respect to the prior work discussing the CS method, the novel aspects of the instant work are then its discussion of the application of CS to peridynamics simulations run in \emph{Peridigm}, a parallel implementation of CS and a compilation of CPU time data, where each of CS, AD, FD and CD are compared.}

@@

An AD system evaluates composition functions in the expected manner: it works by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, observing commonly expected order of operations. However, an AD system does additional work during a function evaluation in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their
partial derivatives may be computed and stored in a sequence. The AD system then multiplies the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. 
%
It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence. This is significant because modifying a single value rather than keeping a list of values corresponding to every level of a composition function, which itself may not be the sole one needing to be evaluated, represents a savings in memory usage, which at large scales is an active concern. In the literature, the AD scheme described here is called \emph{forward automatic differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore in this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, and
particularly \cite{ref-on-AD} which is foundational.

Appendix~\ref{sec:appendixB} includes an example that walks the reader through the process a computer uses to compute derivatives via AD.  Some things to note about AD are that no approximation of derivatives is being made because the analytical forms of the partial derivatives of the elementary math functions are defined alongside them. The accuracy of AD is then limited by the precision of the AD system's definition of the elementary math functions and their partial derivatives.

@@

\note[]{Re: I.5, next section is taken from the thesis.}
\section{Working examples of CS, AD and FD for Newton's method}
\label{sec:WorkingExamples}

@@

\note[]{Re: I.2, John I think you would better handle expanding this section,
I was going to essentially copy silling:psa.}\change[]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics,seleson2014atom}.
